

\begin{itemize}
%{Complex Presentation Issues *** 3 ***}
\item Footnotes and page references (galleys?)
\item Minimal matrix formatting and other advanced formatting
          problems
\item Folding and holophrasting ([1,2,3,4,5,6] =$>$ [..,3,4,..])
\item Graph editing?
\item Ligatures (n things -> 1 pres) (ffi, \alpha, ->, etc.)
\item word breaking (1 thing -> n pres) (chocolaatje, chocola-tje)

\end{itemize}



\par Two clearly distinct forms of editing structured documents exist.
        First, a document can be transformed by giving explicit commands bound to key
        presses or selected from menus. Secondly, after it is freely edited, a presentation can be
        interpreted to yield an updated document. Explicit switching between
        these two forms or modes is a nuisance, especially when the switch can't be
        made because the document is in an incorrect state.
%\section{``Classical'' Structure Editing}

\par The document is edited by performing transformations on the document
        tree. This is a rigid, but in some cases useful way of editing. Edit commands
        can also apply to view state values, like\note{geen like} in a tree browser, when a click on a
        handles switches the expansion state of a node. 
 \begin{itemize}
 
 \item Edit command/transformation specification formalism, like in
          SSL
 \end{itemize}

%\section{Visual Parsing *** 1 ***}

\par Parse an edited presentation tree back into a document tree. 
\par Problems:
 \begin{itemize}
 
 \item Specification formalism
 \item col [row [a,b], row [c,d] ] == row [ col [a,c], col [b,d] ]
 \item Parsing derived values
 \item Parsing duplicate values. Frac Num Den $\{$ pres = num.pres
          `above` num.pres $\}$
 \item Incremental
 \item White Space Handling/ Layout preservation
 \item Invalid Document Handling
 \end{itemize}

 
Do we need a visual parser?
Commands and string parsing enough?
What about rich text paragraph with small formulas, bold text, superscript text, etc. Will look like crap in a string presentation. 
Use commands? ' ' = split word, '$\backslash$ n' = split para? Leads to complex command language and will ultimately be used to create an ad hoc parser.
Some kind of parser required. List separated by ' '?
Maybe not a full blown parser but just a little abstraction.
