% draft?
% 11pt?
% conclusions?


\documentclass[twoside,epsf]{report}
\setcounter{errorcontextlines}{\maxdimen}

\usepackage{thesis}



% dummy labels for chapters until Ch_FormalSpec
%\newlabel{chap:formalSpec}{{5}{91}}
%\newlabel{sect:singleExtra}{{5.2}{97}}
%\newlabel{chap:presenting}{{7}{1}}

% dummy labels for Ch_FormalSpec
\newlabel{chap:requirements}{{2}{13}}
\newlabel{sect:sourceeditor}{{2.1.1}{14}}
\newlabel{sect:editingExtraState}{{2.2.6}{35}}
\newlabel{chap:proxArch}{{3}{49}}
\newlabel{sect:editProces}{{4.1}{77}}
\newlabel{sect:extraState}{{4.2}{79}}
\newlabel{sect:informalDuplicates}{{4.3}{103}}

% dummy labels for Ch_ArchCombs
%\newlabel{chap:proxArch}{{3}{49}}
%\newlabel{sect:archProximaLayers}{{3.3}{58}}
%\newlabel{sect:intrProcess}{{3.5}{66}}
%\newlabel{chap:formalSpec}{{5}{91}}

% dummy labels for Ch_Xprez
%\newlabel{chap:requirements}{{2}{13}}
%\newlabel{sect:presentationFormalism}{{2.2.3}{32}}
%\newlabel{sect:presLevel}{{3.1.3}{53}}
%\newlabel{sect:toolkits}{{2.3.3}{40}}
%\newlabel{sect:ImplPres}{{8.2.4}{194}}

% dummy labels for Ch_Prototype
%\newlabel{chap:requirements}{{2}{13}}
%\newlabel{sect:sourceeditor}{{2.1.1}{14}}
%\newlabel{chap:proxArch}{{3}{49}}
%\newlabel{sect:docLevel}{{3.1.1}{49}}
%\newlabel{sect:archProximaLayers}{{3.3}{56}}
%\newlabel{chap:archCombs}{{6}{139}}
%\newlabel{chap:presenting}{{7}{175}}
%\newlabel{sect:xprezFrac}{{7.2.4}{185}}

% dummy labels for Ch_Conclusions
%\newlabel{chap:requirements}{{2}{13}}
%\newlabel{chap:proxArch}{{3}{49}}
\newlabel{chap:informalSpec}{{4}{77}}
\newlabel{chap:formalSpec}{{5}{91}}
%\newlabel{chap:archCombs}{{6}{139}}
%\newlabel{chap:presenting}{{7}{175}}
%\newlabel{chap:prototype}{{8}{191}}


\begin{document}

\selectlanguage{english}

\pagenumbering{roman}

%\newcommand{\subtitle}{Chapters 7\&8: Prototype and Conclusions}     \include{P_AltTitlePages}
%\include{P_Titlepages}

%\tableofcontents

\clearpage \thispagestyle{empty} \cleardoublepage

\pagenumbering{arabic}

\setcounter{chapter}{0} \include{Ch_Introduction}	% 1
\setcounter{chapter}{1} \include{Ch_XML}   		% 2
\setcounter{chapter}{2} \include{Ch_ProxArch}    	% 3
%\setcounter{chapter}{3} \include{Ch_InformalSpec}	% 4
%\setcounter{chapter}{4} \setcounter{page}{89} \include{Ch_FormalSpec}	% 5
%\setcounter{chapter}{5} \include{Ch_Xprez}   		% 7
\setcounter{chapter}{6} \setcounter{page}{157} \include{Ch_Prototype}    	% 8
\setcounter{chapter}{7} \bc \setcounter{page}{177} \ec \include{Ch_Conclusions}	% 9 yet to be written
 
\bibliographystyle{abbrv}
\bibliography{thesis}

% for chapter with correct bib:
% uncomment "\bibliography.." above 
% tex chapter, bibthesis.bat, tex chapter, texall.bat  (so chap refs are ok), tex chapter
% (biball.bat is only for checking entire thesis for bib problems)

% \include{P_Dankwoord}
% \include{P_Samenvatting}
% \include{P_CV}
%\include{P_IpaDissertations}

%\thispagestyle{empty}
%\cleardoublepage

\end{document}

\bc
IMPORTANT
incrementality vs reuse? check what Proxima really does.

Urgent problems:
!!!figure out reqs and levels in intro and requirements chapters
presentation/rendering image?  mention that lower layers are fixed and hence presentation often used for rendering

General TODOs
* ISBN

* Check for important bits:
* - from CVS: Ch_EditModel.tex, Ch_LayeredArchs.tex, Ch_SingleLayerOld.tex,  and Paper_LayeredArchs.tex  
* - Somewhere.tex, and other old documents

* Extra presentation state and extra interpr. state   versus presentation extra state and interpretation extra state?
* british american  i.e. has , in American english.    is ize possible in british english? If so switch to Br.

* Fix math fonts.

* E elimination vs E introduction

* mention that doc-oriented is always easy. Even for a syn dir editor, the hard bits are in pointing, which is a pres-oriented op.





Before defense:

what happens if mappings are not maintained, what behavior is lost exactly?
bomen in informalspec 1:n etc? why map on nodes instead of subtrees, etc.
Absorption implies CH is HH, even goed uitzoeken hoe dat zit


NOTES
Screen width similar to latex output width:
iii                                                                                                                       

presentation oriented editing is done at layout. 
would be nice to call it presentation. However:
enr.doc --- present -> presentation. level gets name from producing function. 


"higher" is used instead of "upper", because higher level is Level^{High}
upper level is Level_{High} is bit strange, and Level_{Up} does not make sense.








* Order of chapters: layered, single  or single, layered?
* layers, single. PROBLEM edit and diff model must be explained, and is typical single layer stuff
                         GOOD: editing process is already explained before single
                         
* single, layers. PROBLEM assumes edit ops in and out
                                        editing process not explained
                         GOOD: Final bits of layered can give the complete thing, including one layer
                                    edit/diff stuff explained.

single, layers: probably the best choice. edit model is already a bit explained previously. and one of the reasons for edit ops instead of levels is that single layer needs this.
                                    


\ec


