
\documentclass[preprint,natbib]{sigplanconf}

\usepackage{amsmath}
\usepackage{../Utils}

\begin{document}

\conferenceinfo{Haskell Symposium '08}{September 25, 2008, Victoria, British Columbia, Canada.} 
\copyrightyear{2008} 
\copyrightdata{[to be supplied]} 

\titlebanner{Draft, to be submitted to Haskell Symposium '08}        % These are ignored unless
\preprintfooter{\version}   % 'preprint' option specified.

\title{Proxima}
%\subtitle{}

\authorinfo{Martijn M. Schrage \and S. Doaitse Swierstra}
           {Utrecht University}
           {\{martijn,doaitse\}@cs.uu.nl}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
Languages, Human Factors, Documentation

\keywords
keyword1, keyword2




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\bl
\o Proxima
\o what is proxima
\el

\bl
\o Proxima is used to create editor
\el
\bl
\o Contributions:
\o Simple creation of editor for active documents
\o provides easy syncing and static checks, as well as nice UI
\el

This paper ...


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Documenting Bayesian networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o Bayesian networks are ...
\o For creating them, we have Dazzle
\o Dazzle~\cite{schrage05Dazzle}
\o development is complex process
\o During development, choice on arrows, based on expert interviews

\o probtables depend on parents value names, nr of values
\o scripts available for creating initial tables
\o problem is with updates!
\o text speaks about nodes in the network, must be updated when network is changed or renamed
\el



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{An editor for documenting Bayesian networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o Network graph editor. No support for analysis, just construction.
\o import/export of network
\o Views on main network
\o free text editing
\o refs to nodes
\o chapters/sections/subsections and promoting/demoting
\o probability table editing
\el



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proxima}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o Proxima
\o what is proxima
\el
\bl
\o Layered architecture and parameters
\el



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o Presentation
\o Evaluation layer keeps graphs consistent
\o AG for static checks
\o prob tables?
\el

\subsection{Probability tables}

gather 
Scrap your boilerplate.

\subsection{Network views}
problem that not only the views are presented, but also the main network.
Hence, gathering all 



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o no doc editor available
\o why other systems cannot do this
\el


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o manipulation of chapters, etc.
\o Extensions on generic system: menus, find/replace
\el



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bl
\o Future work: extensions required
\el


%\acks


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}

\bibliography{../proxima}

\end{document}

